---
title: "Creating Cohort Subset Definitions"
author: "James P. Gilbert"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: yes
  html_document:
    number_sections: yes
    toc: yes
vignette: >
  %\VignetteIndexEntry{Creating Cohort Subset Definitions}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
options(width = 80)
knitr::opts_chunk$set(
  cache = FALSE,
  comment = "#>",
  error = FALSE)
someFolder <- tempdir()
packageRoot <- tempdir()
baseUrl <- "https://api.ohdsi.org/WebAPI"
```
# Introduction
This guide aims to describe the process of cohort subsetting using `CohortGenerator`.
The purpose of Cohort subsetting operations is to allow the creation of common operations that can be applied to
generated cohorts in order to subset to different operations in a consistent manner.


# Subset defintions
Subset definitions are named sets of operations that can be applied to a set of one or more cohorts.
The current operations that you can apply to cohorts are:

* Limit subsets
* Demographic subsets
* Cohort subsetting

Operations can be sequentially chained within _subset definitions_ and all outputs are considered full cohorts that
can be passed into to other packages as if they are cohorts designed in packages

### Demographic subset operations
This subsetting process allows you to capture the age, race/ethnicity gender within a cohort as subgroups.
For example, "subset cohorts to subjects that are male between the ages 1 and 5 years old".

### Subsetting to other cohorts
This type of operation allows you to subset a cohort to only those subjects included in one or more other cohorts


# Creating Cohort subset definition
First get a Cohort definition set:
```{r results='hide', error=FALSE, warning=FALSE, message=FALSE}
library(CohortGenerator)
# A list of cohort IDs for use in this vignette
cohortIds <- c(1778211, 1778212, 1778213)
# Get the SQL/JSON for the cohorts
cohortDefinitionSet <- ROhdsiWebApi::exportCohortDefinitionSet(baseUrl = baseUrl,
                                                               cohortIds = cohortIds)
```

A definition can include different subset operations - these are applied strictly in order:

```{r results='hide', error=FALSE, warning=FALSE, message=FALSE}
# Example, we want to have a HTN cohort that starts any time prior to the index start
# and the HTN cohort ends any time after the index start
subsetDef <- createCohortSubsetDefinition(
  name = "Patients in cohort cohort 1778213 with 365 days prior observation",
  definitionId = 1,
  subsetOperators = list(
    # here we are saying 'first subset to only those patients in cohort 1778213'
    createCohortSubset(id = 1001,
                       name = "Subset to patients in cohort 1778213",
                       # Note that this can be set to any id - if the cohort is empty or doesn't exist this will not error
                       cohortIds = 1778213,
                       cohortCombinationOperator = "any",
                       negate = FALSE,
                       startWindow = createSubsetCohortWindow(startDay = -9999,
                                                              endDay = 0,
                                                              targetAnchor = "cohortStart"),
                       endWindow = createSubsetCohortWindow(startDay = 0,
                                                            endDay = 9999,
                                                            targetAnchor = "cohortStart")),

    # Next, subset to only those with 365 days of prior observation
    createLimitSubset(id = 1002,
                      name = "Observation of at least 365 days prior",
                      priorTime = 365,
                      followUpTime = 0,
                      limitTo = "firstEver")
  )
)
```

Next we create a similar definition that also subsets the specified cohorts to require patients with specific
demographic criteria.
We can do that by copying the subset operations from our first definition and modifying them.
```{r results='hide', error=FALSE, warning=FALSE, message=FALSE}
subsetOperations2 <- subsetDef$subsets

# subset to those between aged 18 an 64
subsetOperations[[3]] <-
  createDemographicSubset(id = 1003,
                          name = "18 - 65",
                          ageMin = 18,
                          ageMax = 64)

subsetDef2 <- createCohortSubsetDefinition(
  name = "Patients in cohort 1778213 with 365 days prior observation, aged 18 - 64",
  definitionId = 2,
  subsetOperators = subsetOperations2)
```
Next we need to add the subset definitions to the base cohort set.
This will automatically add identifiers and OHDSI SQL for the subset cohorts as well as storing references
for saving definition sets for re-use.
```{r results='hide', error=FALSE, warning=FALSE, message=FALSE}
cohortDefinitionSet <- cohortDefinitionSet |>
  addCohortSubsetDefinition(subsetDef) |>
  addCohortSubsetDefinition(subsetDef2)
```
Note that when adding a subset definition to a cohort definition set, the target cohort ids e.g (1778211, 1778212) must
exist in the `cohortDefinitionSet` and the output ids (`1778211002`, `1778212003`) must be unique.
As with all cohorts, any cohorts with these ids will be deleted prior to execution to prevent collisions.
Note that the default expression for output cohort ids is `targetId * 1000 + definitionId` this may cause collisions
that will cause `addCohortSubsetDefinition` to error.

### Generating subsets
Executing CohortGenerator, we can now include the subset operations when our cohorts are generated:

```{r results='hide', error=FALSE, warning=FALSE, message=FALSE, eval = FALSE}

connectionDetails <- Eunomia::getEunomiaConnectionDetails()
createCohortTables(connectionDetails = connectionDetails,
                   cohortDatabaseSchema = "main",
                   cohortTableNames = getCohortTableNames("my_cohort"))
# ### As subsets are a big side effect we need to be clear what was generated and have good naming conventions
generatedCohorts <- generateCohortSet(connectionDetails = connectionDetails,
                                      cdmDatabaseSchema = "main",
                                      cohortDatabaseSchema = "main",
                                      cohortTableNames = getCohortTableNames("my_cohort"),
                                      cohortDefinitionSet = cohortDefinitionSet,
                                      incremental = TRUE,
                                      incrementalFolder = file.path(someFolder, "RecordKeeping"))
```
Cohort subset definitions can be run incrementally.
In fact, if the base cohort definition changes for any reason, any subsets will automatically be re-executed when calling
`generateCohortSet`.

# Saving and loading subset definitions
Subset definitions can be converted to JSON objects as follows:
```{r results='hide', eval=FALSE}
jsonDefinition <- subsetDef$toJSON()
```

For the purpose of writing to disk we recommend the use of `ParallelLogger` for consistency.
```{r results='hide', eval=FALSE}
# Save to a file
ParallelLogger::saveSettingsToJson(subsetDef$toList(), "subsetDefinition1.json")
```
