---
title: "Creating Cohort Subset Definitions"
author: "James P. Gilbert"
date: "`r Sys.Date()`"
output:
pdf_document:
toc: yes
html_document:
number_sections: yes
toc: yes
vignette: >
%\VignetteIndexEntry{Creating Cohort Subset Definitions}
%\VignetteEncoding{UTF-8}
%\VignetteEngine{knitr::rmarkdown}
editor_options:
chunk_output_type: console
---

```{r setup, include=FALSE}
options(width = 80)
knitr::opts_chunk$set(
  cache = FALSE,
  comment = "#>",
  error = FALSE)
someFolder <- tempdir()
packageRoot <- tempdir()
baseUrl <- "https://api.ohdsi.org/WebAPI"
```
# Creating Cohort subset definition
First get a Cohort definition set:
```{r results='hide', error=FALSE, warning=FALSE, message=FALSE}
library(CohortGenerator)
# A list of cohort IDs for use in this vignette
cohortIds <- c(1778211, 1778212, 1778213)
# Get the SQL/JSON for the cohorts
cohortDefinitionSet <- ROhdsiWebApi::exportCohortDefinitionSet(baseUrl = baseUrl,
                                                               cohortIds = cohortIds)
```

A definition can include different subset operations - these are applied strictly in order:

```{r results='hide', error=FALSE, warning=FALSE, message=FALSE}
# Example, we want to have a HTN cohort that starts any time prior to the index start
# and the HTN cohort ends any time after the index start
subsetOperations <- list(
  # here we are saying 'first subset to only those patients in cohort 1778213'
  createCohortSubset(id = 1001,
                     name = "Subset to patients in cohort 1778213",
                     cohortIds = 1778213,
                     cohortCombinationOperator = "any",
                     negate = FALSE,
                     startWindow = createSubsetCohortWindow(startDay = -9999,
                                                            endDay = 0,
                                                            targetAnchor = "cohortStart"),
                     endWindow = createSubsetCohortWindow(startDay = 0,
                                                          endDay = 9999,
                                                          targetAnchor = "cohortStart")),
  
  # Next, subset to only those with 365 days of prior observation
  createLimitSubset(id = 1002,
                    name = "Observation of at least 365 days prior",
                    priorTime = 365,
                    followUpTime = 0,
                    limitTo = "firstEver")
)
```
We then need to capture these definitions in a definition object

```{r results='hide', error=FALSE, warning=FALSE, message=FALSE}
subsetDef <- createCohortSubsetDefinition(name = "Patients in cohort 1778213 with 365 days prior observation",
                                          definitionId = 1,
                                          targetOutputPairs = list(c(1778211, 1778211001), c(1778212, 1778212001)),
                                          subsets = subsetOperations)
```

Next we create a similar definition that also subsets the specified cohorts to require patients with specific
demographic criteria:

```{r results='hide', error=FALSE, warning=FALSE, message=FALSE}
subsetOperations2 <- subsetOperations

# subset to those between aged 18 an 64
subsetOperations2[[3]] <- createDemographicSubset(id = 1003,
                                                  name = "18 - 65",
                                                  ageMin = 18,
                                                  ageMax = 64)

subsetDef2 <- createCohortSubsetDefinition(name = "Patients in cohort 1778213 with 365 days prior observation, aged 18 - 64",
                                           definitionId = 2,
                                           # NOTE: The output ids should not clash with any that already exist
                                           targetOutputPairs = list(c(1778211, 1778211002), c(1778212, 1778212003)),
                                           subsets = subsetOperations2)

cohortDefinitionSet <- cohortDefinitionSet |>
      addCohortSubsetDefinition(subsetDef) |>
      addCohortSubsetDefinition(subsetDef2)
```
Note that when adding a subset definition to a cohort definition set, the target cohort ids e.g (1778211, 1778212) must
exist in the `cohortDefinitionSet` and the output ids (`1778211002`, `1778212003`) must be unique.
As with all cohorts, any cohorts with these ids will be deleted upon execution.

### Generating subsets
Executing CohortGenerator, we can now include the subset operations when our cohorts are generated:

```{r results='hide', error=FALSE, warning=FALSE, message=FALSE, eval = FALSE}

connectionDetails <- Eunomia::getEunomiaConnectionDetails()
createCohortTables(connectionDetails = connectionDetails, 
                   cohortDatabaseSchema = "main",
                   cohortTableNames = getCohortTableNames("my_cohort"))
# ### As subsets are a big side effect we need to be clear what was generated and have good naming conventions
generatedCohorts <- generateCohortSet(connectionDetails = connectionDetails,
                                      cdmDatabaseSchema = "main",
                                      cohortDatabaseSchema = "main",
                                      cohortTableNames = getCohortTableNames("my_cohort"),
                                      cohortDefinitionSet = cohortDefinitionSet,
                                      incremental = TRUE,
                                      incrementalFolder = file.path(someFolder, "RecordKeeping"))
```
Cohort subset definitions can be run incrementally.
In fact, if the base cohort definition changes for any reason, any subsets will automatically be re-executed when calling
`generateCohortSet`.

# Saving subset definitions
Subset definitions can be converted to JSON objects as follows:
```{r results='hide', eval=FALSE}
jsonDefinition <- subsetDef$toJSON()
```

For the purpose of writing to disk we recommend the use of `ParallelLogger` for consistency.
```{r results='hide', eval=FALSE}
# Save to a file
ParallelLogger::saveSettingsToJson(subsetDef$toList(), "subsetDefinition1.json")
```
